INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-11-09 chapters/background.tex
small changes

The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
In particular, language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.

>>>>

The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
In particular, language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
To compare, the goals behind our \textsc{Ingrid} approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.

### 2016-11-09 chapters/evaluation.tex
small changes

We implemented the proposed INGRID method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call \PV{nebo "to bind with"} the MPS API,  \PV{nebo "which was used to programatically generate elements of the new language} e.g. in order to programmatically create new language concepts and their aspects.
The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
Our complete implementation is available at \url{https://github.com/premun/ingrid}.

For the purpose of evaluation, we applied INGRID to several well-established and widely used languages, including JSON, JavaScript, and C\#.
\PV{Dalsi vetu bych odebral a treba jen v predchozi vete dal JavaScript (ECMAScript 5.1). Usetri to hodne mista a informace zustane stejna.}
We used the version of JavaScript that is standardized as ECMAScript 5.1~\cite{ref:ECMASCRIPT51} --- the specification was finalized in June 2011, and it is currently the most frequently adopted version.
MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
In the rest of this section, we discuss our experience with application of INGRID to these languages, and then we highlight few general observations.

However, first we must emphasize that MPS languages automatically produced by the INGRID method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
\PV{Nejsem si jisty, jestli je to vubec mozne. To, ze nejake pravidlo "ma typ", je asi az moc high-level informace. To, ze jedno pravidlo je definice metody a jine je volani metody, take z gramatiky nikdy ziskat nepujde. Bud vetu odstranit nebo preformulovat (bez yet atd.)}
INGRID does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
\PV{Toto take neni pravda - editor je v kazdem jazyku lehce upraveny (to je tech 20 min / 1 hodina ...). Structure aspect je stejny, textgen je take krome Csharpu upraveny.}
For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by INGRID and (2) the adjusted ANTLR v4 grammar used as input for INGRID, both in the repository that contains also our implementation.

@@

The main advantage of INGRID over ECMAScript4MPS is that, despite its current limitations, INGRID fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
INGRID achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
The Structure aspect generated by INGRID is very similar to that of the ECMAScript4MPS project, which was created manually \PV{over misto in?} in a large number of hours.

\paragraph{C\#.}
Another very complex programming language that we used for experiments is C\#.
Before we could run INGRID, we had to manually adjust the ANTLR grammar of C\# in order to ensure that INGRID produces a reasonably useful MPS language.
\PV{Na gramatiku asi stacilo 20 minut, kdyz clovek vi, co dela. 40 minut pak na castecnou upravu editoru.}
Roughly one hour of manual effort was needed for this step.
Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.

@@

\paragraph{Other languages.}
We also tried to apply INGRID on few other languages, such as Python and Ruby.
Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with INGRID.
\PV{Tim not compatible je spis myslena hiearchizace gramatiky a styl rozdeleni pravidel. Byl trochu jiny nez jinde a dochazelo tam k dost zmatenym strukturam (co se MPS tyce). Cyklicka pravidla jsou asi nejmensi problem, daji se odstranit prakticky hned a vetsinou celkem bez ujmy.}
For example, INGRID cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
Significant modifications of the grammar would be necessary.
\PV{Mozna to ani neni jak adresovat, proste se bude muset upravit gramatika.}
We plan to address this limitation in the future.

\paragraph{General observations.}
The main overall benefit of the INGRID method is partial automation.
Most of the languages discussed above, which we tried to create in MPS by using INGRID, are quite complex regarding their structure and syntax variety.
Therefore, completely manual definition would be a very time-consuming and error-prone process.
Fully automated generation of the Structure aspect is where INGRID spares the user from many hours of tedious and sometimes quite challenging work.

On the other hand, our experiments with complex languages show that, in the case of the Editor and Textgen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast.
All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
Despite that, we also tried to design some automated heuristics during our work on the INGRID project, but so far all yield rather average results especially when compared to what human users can achieve efficiently \PV{instead}.

>>>>

We implemented the proposed \textsc{Ingrid} method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call \PV{nebo "to bind with"} the MPS API,  \PV{nebo "which was used to programatically generate elements of the new language} e.g. in order to programmatically create new language concepts and their aspects.
The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
Our complete implementation is available at \url{https://github.com/premun/ingrid}.

For the purpose of evaluation, we applied \textsc{Ingrid} to several well-established and widely used languages, including JSON, JavaScript, and C\#.
\PV{Dalsi vetu bych odebral a treba jen v predchozi vete dal JavaScript (ECMAScript 5.1). Usetri to hodne mista a informace zustane stejna.}
We used the version of JavaScript that is standardized as ECMAScript 5.1~\cite{ref:ECMASCRIPT51} --- the specification was finalized in June 2011, and it is currently the most frequently adopted version.
MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
In the rest of this section, we discuss our experience with application of \textsc{Ingrid} to these languages, and then we highlight few general observations.

However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
\PV{Nejsem si jisty, jestli je to vubec mozne. To, ze nejake pravidlo "ma typ", je asi az moc high-level informace. To, ze jedno pravidlo je definice metody a jine je volani metody, take z gramatiky nikdy ziskat nepujde. Bud vetu odstranit nebo preformulovat (bez yet atd.)}
\textsc{Ingrid} does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
\PV{Toto take neni pravda - editor je v kazdem jazyku lehce upraveny (to je tech 20 min / 1 hodina ...). Structure aspect je stejny, textgen je take krome Csharpu upraveny.}
For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by \textsc{Ingrid} and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}, both in the repository that contains also our implementation.

@@

The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
\textsc{Ingrid} achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually \PV{over misto in?} in a large number of hours.

\paragraph{C\#.}
Another very complex programming language that we used for experiments is C\#.
Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces a reasonably useful MPS language.
\PV{Na gramatiku asi stacilo 20 minut, kdyz clovek vi, co dela. 40 minut pak na castecnou upravu editoru.}
Roughly one hour of manual effort was needed for this step.
Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.

@@

\paragraph{Other languages.}
We also tried to apply \textsc{Ingrid} on few other languages, such as Python and Ruby.
Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.
\PV{Tim not compatible je spis myslena hiearchizace gramatiky a styl rozdeleni pravidel. Byl trochu jiny nez jinde a dochazelo tam k dost zmatenym strukturam (co se MPS tyce). Cyklicka pravidla jsou asi nejmensi problem, daji se odstranit prakticky hned a vetsinou celkem bez ujmy.}
For example, \textsc{Ingrid} cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
Significant modifications of the grammar would be necessary.
\PV{Mozna to ani neni jak adresovat, proste se bude muset upravit gramatika.}
We plan to address this limitation in the future.

\paragraph{General observations.}
The main overall benefit of the \textsc{Ingrid} method is partial automation.
Most of the languages discussed above, which we tried to create in MPS by using \textsc{Ingrid}, are quite complex regarding their structure and syntax variety.
Therefore, completely manual definition would be a very time-consuming and error-prone process.
Fully automated generation of the Structure aspect is where \textsc{Ingrid} spares the user from many hours of tedious and sometimes quite challenging work.

On the other hand, our experiments with complex languages show that, in the case of the Editor and Textgen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast.
All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
Despite that, we also tried to design some automated heuristics during our work on the \textsc{Ingrid} project, but so far all yield rather average results especially when compared to what human users can achieve efficiently \PV{instead}.

### 2016-11-09 chapters/introduction.tex
small changes

In this paper we present INGRID, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
We discuss the main challenges that we encountered during the development of INGRID and our solution to them.
Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
We provide more details in Section~\ref{sect:LANGDEF}.

The INGRID tool can help also to make the construction of DSLs more efficient.
Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS using INGRID, than to create the language manually using the MPS GUI.

%INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.

The rest of the paper has the following structure.
In the next section, we provide necessary information about MPS and a brief overview of projects with similar goals as INGRID.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
We also discuss our experience with usage of INGRID on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

>>>>

In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
We provide more details in Section~\ref{sect:LANGDEF}.

The \textsc{Ingrid} tool can help also to make the construction of DSLs more efficient.
Based on our experience, we believe that, especially for simple languages, it is less time consuming to write their grammar by hand and then create the language definition in MPS using \textsc{Ingrid}, than to create the language manually using the MPS GUI.

%\textsc{Ingrid} allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.

The rest of the paper has the following structure.
In the next section, we provide necessary information about MPS and a brief overview of projects with similar goals as \textsc{Ingrid}.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

### 2016-11-09 chapters/language_definition.tex
small changes

The proposed INGRID method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

The process of MPS language construction, as performed by INGRID, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen).
Each of the phases 2-4 is exclusively and fully responsible for one aspect.

INGRID is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by INGRID.

@@

We will describe the INGRID method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.

@@

The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST are performed in this phase.
An output is a custom AST-like structure representing the grammar, especially the hierarchy of parser rules.

@@

As a part of the INGRID method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
However, before focusing on the intermediate layers and the shortcut approach, we describe the basic principles of translation from the AST into the language structure.

@@

Our solution that we implemented in the current version of the INGRID method is to create all the cells and place them in a single row.
The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently in the MPS IDE.
The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.

@@

Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
We expect that a typical user of the INGRID method will be able to use the projectional editor in MPS quite efficiently.
More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.

@@

Our INGRID method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
An obvious choice would therefore be to use the same approach for Editor and TextGen.
Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.

@@

During our work on the INGRID method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
We discuss several issues and our workarounds in this section.

\paragraph{Adjusting grammars.}
In some cases, a small adjustment of the input grammar before the run of INGRID might yield a better and more useful MPS language.
Here we illustrate this on two simple examples.

@@

\paragraph{Breaking original grammars and parsers.}
We also want to point out a general problem with grammar adjustments that we encountered, and which all potential users of the INGRID method should be aware of.
The original ANTLR grammar for an input language can be very easily changed in a way that, at first, seems harmless and valid inside MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
In particular, adjustments needed to improve the resulting MPS language can break down the parser.

This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose (that was not expected) in our project.
Since, for practical reasons, it is very important that a parser works even for documents (programs) written according to the original grammar, adjustments have to be performed very carefully by the users of INGRID.

>>>>

The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

The process of MPS language construction, as performed by \textsc{Ingrid}, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects of the language (Structure, Editor, and TextGen).
Each of the phases 2-4 is exclusively and fully responsible for one aspect.

\textsc{Ingrid} is currently able to create (i) a full Structure aspect for each element (type of an AST node) of the given language, (ii) a very basic Editor, and (iii) a basic TextGen aspect.
Therefore, the resulting MPS language typically still has to be adjusted manually to improve its usability, and the human users must also define the remaining aspects not yet supported by \textsc{Ingrid}.

@@

We will describe the \textsc{Ingrid} method using the example of a simplified XML language that is defined in Figure~\ref{fig:SIMPLEXML}.
The \emph{SimpleXML} language is small but complex enough to be used for illustration of the main challenges and behavior of the proposed algorithms.

@@

The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the \textsc{Ingrid} algorithm.
In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST are performed in this phase.
An output is a custom AST-like structure representing the grammar, especially the hierarchy of parser rules.

@@

As a part of the \textsc{Ingrid} method, we have designed an approach to eliminate the unnecessary layers during construction of the Structure aspect --- we call it the \emph{shortcut} approach (Section~\ref{sect:SHORTCUT}).
However, before focusing on the intermediate layers and the shortcut approach, we describe the basic principles of translation from the AST into the language structure.

@@

Our solution that we implemented in the current version of the \textsc{Ingrid} method is to create all the cells and place them in a single row.
The resulting basic layout is illustrated on the example of the \mpsconcept{Element{\_}1} concept in the upper left corner of Figure~\ref{fig:EDITORADJUST}.
Further adjustments of the layout, such as indentation and line breaks, can be done very efficiently in the MPS IDE.
The bottom right part of Figure~\ref{fig:EDITORADJUST} shows a fully customized layout for the \mpsconcept{Element{\_}1} concept.

@@

Based on our experience, it takes a very short time to manually adjust the layout into a form much better than any fully automated heuristic could achieve.
We expect that a typical user of the \textsc{Ingrid} method will be able to use the projectional editor in MPS quite efficiently.
More details are provided in Section~\ref{sect:EVAL}, including our experience with several mainstream programming languages.

@@

Our \textsc{Ingrid} method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
An obvious choice would therefore be to use the same approach for Editor and TextGen.
Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.

@@

During our work on the \textsc{Ingrid} method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
We discuss several issues and our workarounds in this section.

\paragraph{Adjusting grammars.}
In some cases, a small adjustment of the input grammar before the run of \textsc{Ingrid} might yield a better and more useful MPS language.
Here we illustrate this on two simple examples.

@@

\paragraph{Breaking original grammars and parsers.}
We also want to point out a general problem with grammar adjustments that we encountered, and which all potential users of the \textsc{Ingrid} method should be aware of.
The original ANTLR grammar for an input language can be very easily changed in a way that, at first, seems harmless and valid inside MPS, but the parser generated out of the adjusted grammar stops accepting the original language.
In particular, adjustments needed to improve the resulting MPS language can break down the parser.

This problem has two main causes: (1) low-level implementation details of token matching in the ANTLR parser and (2) usage of ANTLR grammars for a different purpose (that was not expected) in our project.
Since, for practical reasons, it is very important that a parser works even for documents (programs) written according to the original grammar, adjustments have to be performed very carefully by the users of \textsc{Ingrid}.
